---
title: "EXAM 1"
author: "Jules Pellin"
date: "2026-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

R has five main data structures. First, we have the **Vectors**, they are one-dimensional sequences where all elements must be the same type, which means that they are the building block of every other structure in R. Then we have the **Matrices**, they are two-dimensional and also require all elements to be the same type, making them the most efficient for numerical operations. **Arrays**, they extend matrices to more than two dimensions but are rarely used in standard data analysis. The **Lists** are the most generic structure because they have no type restrictions, meaning each element can be anything, including other lists, vectors, or matrices. This makes them ideal for storing heterogeneous or hierarchical data. The **Data frames** are the most useful structure for data analysis, since they are two-dimensional like a matrix, but each column can hold a different type. For example, when you import a CSV with read.csv(), you get a data frame. In practice, vectors are used for single variables, matrices for uniform numerical data, lists when data is mixed or nested, and data frames for any tabular dataset.

## Question 2

### Q2a & b
For Q2a, I used list.files() to iterate through all subfolders and read 
each CSV file into R, then combined them into a single data frame using 
rbind(). For Q2b, I extracted the county name, crop name, latitude, and 
longitude directly from the file path structure and added them as four 
new columns to the merged data frame.

```{r}
# Set base path
base_path <- "C:/Users/pelli/Downloads/CropModelResults/CropModelResults"

# Get all CSV file paths with full absolute paths
all_files <- list.files(base_path, recursive = TRUE, 
                        full.names = TRUE, pattern = "seasonal_result\\.csv$")

# Read all files into a list, adding metadata columns
all_data_list <- lapply(all_files, function(f) {
  temp  <- read.csv(f)
  parts <- strsplit(f, "/")[[1]]
  n     <- length(parts)
  
  # Extract county, crop, and lat/lon from path
  county_name   <- parts[n-3]
  crop_name     <- parts[n-2]
  latlon_str    <- parts[n-1]
  lat_lon_split <- strsplit(latlon_str, "N")[[1]]
  
  temp$county    <- county_name
  temp$crop      <- crop_name
  temp$latitude  <- as.numeric(lat_lon_split[1])
  temp$longitude <- -as.numeric(gsub("W", "", lat_lon_split[2]))
  temp
})

# Combine all into one data frame
merged_df <- do.call(rbind, all_data_list)

str(merged_df)
head(merged_df)
```

## Q2c

For Q2c, I renamed the columns "irrig" to "irrigation_demand" and "precip" 
to "precipitation" using colnames(). I then exported the merged data frame 
as a CSV file using write.csv().
```{r}
# ── Q2c: Rename columns and export as CSV ──────────────────────────

# Rename irrig to irrigation_demand and precip to precipitation
colnames(merged_df)[colnames(merged_df) == "irrig"]  <- "irrigation_demand"
colnames(merged_df)[colnames(merged_df) == "precip"] <- "precipitation"

# Verify the rename worked
colnames(merged_df)

# Export as CSV
write.csv(merged_df, "C:/Users/pelli/Documents/econ396/merged_crop_data.csv", row.names = FALSE)
```

###Q2d

For Q2d, I used aggregate() to summarize the total annual irrigation 
demand grouped by crop name and county name.
```{r}
# ── Q2d: Summarize annual irrigation demand by crop and county ──────

# Aggregate total irrigation demand by crop and county
irrigation_summary <- aggregate(irrigation_demand ~ crop + county, 
                                data = merged_df, 
                                FUN = sum)

# Print the summary
print(irrigation_summary)
```

###Q2e

For Q2e, I extracted the year from the date column, then subsetted the 
data for Winter Wheat in Walla Walla at the specific coordinates 
46.03125N, -118.40625W, and calculated the average yield for three 
different year ranges.
```{r}
# ── Q2e: Average yield of Winter Wheat in Walla Walla ──────────────

# Extract year from the date column
merged_df$year <- as.integer(substr(merged_df$YYYY.MM.DD.DOY., 1, 4))

# Subset for the specific location
ww <- merged_df[merged_df$county == "WallaWalla" & 
                merged_df$crop == "Winter_Wheat" & 
                merged_df$latitude == 46.03125 & 
                merged_df$longitude == -118.40625, ]

# Average yield for each year range
mean_1981_1990 <- mean(ww$yield[ww$year >= 1981 & ww$year <= 1990])
mean_1991_2000 <- mean(ww$yield[ww$year >= 1991 & ww$year <= 2000])
mean_2001_2019 <- mean(ww$yield[ww$year >= 2001 & ww$year <= 2019])

print(mean_1981_1990)
print(mean_1991_2000)
print(mean_2001_2019)
```

###Q2f

For Q2f, I subsetted the data for Corn grain between 2001 and 2019, 
then calculated the average yield per location using aggregate(), 
and identified the location with the highest average yield.
```{r}
# ── Q2f: Highest average yield for Corn grain (2001-2019) ──────────

# Subset for Corn grain between 2001 and 2019
corn <- merged_df[merged_df$crop == "Corn_grain" & 
                  merged_df$year >= 2001 & 
                  merged_df$year <= 2019, ]

# Average yield by location
corn_avg <- aggregate(yield ~ county + latitude + longitude, 
                      data = corn, 
                      FUN = mean)

# Print all averages
print(corn_avg)

# Identify the highest yield location
corn_avg[which.max(corn_avg$yield), ]
```

## Question 3

Honestly, the data provided was not well described. I found a lot of missing information that includes units for the variables (e.g. is yield in kg/ha? is irrigation in mm?), a description of what each column represents, and the source and model used to generate the data.

As best practice, metadata should include a clear description of each 
variable and its units, the data source and collection method, spatial 
and temporal coverage, and any limitations of the data.

