---
title: "Final Project"
author: "Jules Pellin"
date: "2026-06-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective

This project analyzes the physical profiles of Olympic athletes across sports, 
sexes, and time using 120 years of Olympic history data (1896–2016). The central 
question is: how do height, weight, and BMI differ across sports, and how has the 
average age of Olympic athletes changed over time? As a former competitive handball 
player, I have a personal interest in understanding how athlete body types vary 
across different disciplines.

## Data

The dataset comes from Kaggle and contains two files: athlete_events.csv with 
271,116 observations across 15 variables (name, sex, age, height, weight, sport, 
event, medal, year, country code), and noc_regions.csv which maps country codes 
to region names. Each row represents one athlete's participation in one Olympic 
event, making it tidy by definition, every column is a variable, every row is 
an observation, and every cell contains a single value.

## Workflow

The script starts by loading and joining both files, then narrows things down to 
Summer Olympics only and drops athletes with missing height or weight data. A BMI 
column gets added from there, and the rest of the project builds off that cleaned 
dataset through four visualizations exploring how physical profiles vary across 
sports, sexes, and time.

```{r load, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggridges)

# Load both files
athletes <- read.csv("athlete_events.csv")
noc      <- read.csv("noc_regions.csv")

head(athletes)
str(athletes)
```

```{r clean, message=FALSE, warning=FALSE}
# Join athlete data with region names by NOC code
dat <- merge(athletes, noc, by = "NOC", all.x = TRUE)

# Filter to Summer Olympics only
dat <- dat[dat$Season == "Summer", ]

# Remove rows where Height or Weight are NA (needed for physical analysis)
dat_clean <- dat[!is.na(dat$Height) & !is.na(dat$Weight), ]

# Calculate BMI
dat_clean$BMI <- dat_clean$Weight / ((dat_clean$Height / 100)^2)

# Report missing data handling
cat("Original rows:", nrow(athletes), "\n")
cat("After Summer filter:", nrow(dat), "\n")
cat("After removing NA Height/Weight:", nrow(dat_clean), "\n")

head(dat_clean[, c("Name", "Sport", "Height", "Weight", "BMI", "region")])
```


## Exploratory Analysis

### Visualization 1 - Height distribution by sport (Ridge Plot)

```{r viz1-ridge, message=FALSE, warning=FALSE}
# Select top 10 sports by athlete count for readability
top_sports <- dat_clean %>%
  count(Sport) %>%
  arrange(desc(n)) %>%
  slice_head(n = 10) %>%
  pull(Sport)

ridge_dat <- dat_clean[dat_clean$Sport %in% top_sports, ]

ggplot(ridge_dat, aes(x = Height, y = Sport, fill = Sport)) +
  geom_density_ridges() +
  labs(title = "Height Distribution Across Top 10 Olympic Sports",
       x = "Height (cm)",
       y = "Sport") +
  theme_minimal() +
  theme(legend.position = "none")
```

Height differences across sports are pretty striking. Gymnastics immediately jumps 
out with its distribution sitting way to the left, shorter athletes just have a 
natural advantage there. Wrestling has this interesting double-hump shape which 
makes sense since it has so many different weight and size categories. Swimming, 
Rowing, and Sailing all lean toward the taller end around 180-185 cm, while 
Athletics has by far the widest spread, which you'd expect from a category that 
lumps marathon runners and shot putters together.


### Visualization 2 - Height vs Weight by Sex

```{r viz2-scatter, message=FALSE, warning=FALSE}
# Sample 3000 rows for readability
set.seed(42)
sample_dat <- dat_clean[sample(nrow(dat_clean), 3000), ]

ggplot(sample_dat, aes(x = Height, y = Weight, color = Sex)) +
  geom_point(alpha = 0.4, size = 1.5) +
  labs(title = "Height vs Weight of Olympic Athletes by Sex",
       x = "Height (cm)",
       y = "Weight (kg)") +
  theme_minimal()
```


There's a clear positive relationship between height and weight across all athletes. 
Male athletes are consistently taller and heavier, with their cluster sitting notably 
higher and to the right of the female cluster. Both sexes follow the same general 
linear trend though, taller athletes weigh more regardless of sex. The few extreme 
outliers at the top are likely weightlifters or shot putters where raw mass is an 
advantage.



### Visualization 3 - BMI by Sport

```{r viz3-bmi, message=FALSE, warning=FALSE}
ggplot(ridge_dat, aes(x = Sport, y = BMI, fill = Sport)) +
  geom_boxplot() +
  labs(title = "BMI Distribution Across Top 10 Olympic Sports",
       x = "Sport",
       y = "BMI") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

Gymnastics has the lowest median BMI of all sports, which makes sense given the 
emphasis on strength-to-weight ratio and body control. Wrestling sits at the 
highest end which is expected since heavier body mass can be an advantage and 
the sport has multiple weight categories. Athletics has the most outliers by far, 
again reflecting how diverse the events within it are. Most sports cluster between 
a BMI of 22 and 26 which falls within the normal to athletic range.



### Visualization 4 - Average athlete age over time

```{r viz4-age, message=FALSE, warning=FALSE}
age_trend <- dat_clean %>%
  group_by(Year) %>%
  summarise(avg_age = mean(Age, na.rm = TRUE))

ggplot(age_trend, aes(x = Year, y = avg_age)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(title = "Average Age of Olympic Athletes Over Time",
       x = "Year",
       y = "Average Age") +
  theme_minimal()
```

Average age has been all over the place historically, which makes sense given how 
different the early Olympics were from today. The spike around 1948 is interesting,
that's likely athletes whose careers got interrupted by WWII finally getting their 
shot. From the 1950s onward it dips as younger athletes started dominating, then 
climbs steadily from 1980 to about 26.5 in 2016. That upward trend probably reflects 
better sports science and recovery letting athletes compete at a high level for longer.


## Reflection

The biggest thing I learned from this project is how much cleaning and filtering 
matters before you even think about visualizing. With 271,000 rows and over 60,000 
NAs in Height and Weight alone, jumping straight into plots would have been complicated to organize. 
Dropping the NA rows rather than filling them made more sense here since the whole 
point was to analyze physical characteristics accurately. The ggridges package ended 
up being one of my favorite parts, showing height distributions across 10 sports 
in one clean visual would have taken way more plots otherwise. Overall I feel like 
the project answered what I set out to explore: athlete body types genuinely differ 
across sports, and Olympic athletes are getting older over time.