---
title: "Homework 5"
author: "Jules Pellin"
date: "2026-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Question 1

```{r q1-load, message=FALSE, warning=FALSE}
# Load required packages
library(readxl)
library(tidyverse)

# Read COVID and demographic data (files are Excel format despite .csv extension)
covid <- read_excel("US-cnty-COVID.csv")
demo  <- read_excel("US-cnty-demo.csv")

# Examine both files
head(covid)
head(demo)
str(covid)
str(demo)
```
```{r q1-merge}
# Merge by BOTH county AND state to avoid duplicate county name bug
# (e.g. King County exists in both WA and TX)
merged <- merge(covid, demo,
                by.x = c("County", "State"),
                by.y = c("County.name", "ST.Name"))

# Verify row count — should still be 3,142
nrow(merged)

# Create rate columns
merged$cases_per_pop    <- merged$Cases  / merged$TotalPop
merged$deaths_per_pop   <- merged$Deaths / merged$TotalPop
merged$deaths_per_senior <- merged$Deaths / merged$Pop.Over65

head(merged)
```
```{r q1-arrange}
# i) Arrange by cases per population (highest first)
cases_ranked <- arrange(merged, desc(cases_per_pop))
head(cases_ranked[, c("County", "State", "cases_per_pop")])

# ii) Arrange by deaths per population (highest first)
deaths_ranked <- arrange(merged, desc(deaths_per_pop))
head(deaths_ranked[, c("County", "State", "deaths_per_pop")])

# iii) Arrange by deaths per senior population (highest first)
# Exclude counties where Pop.Over65 = 0 (division by zero gives Inf)
senior_ranked <- arrange(merged, desc(deaths_per_senior))
senior_ranked_clean <- senior_ranked[is.finite(senior_ranked$deaths_per_senior), ]
head(senior_ranked_clean[, c("County", "State", "deaths_per_senior")])
```
The three rankings reveal very different patterns of COVID burden across US counties.

Ranked by cases per population, small counties with large institutional populations top the 
list: prisons (Trousdale, TN; Norton, KS) and military bases (Chattahoochee, GA) explain 
the extreme rates, as these dense enclosed environments drive rapid transmission regardless 
of county size.

Ranked by deaths per population, a different set of small rural counties appears (Gove, KS; 
Jerauld, SD; Hancock, GA), likely reflecting older and more vulnerable local populations 
where fatality rates ran high even with fewer total cases.

The senior population ranking required filtering out counties with zero seniors recorded, 
which produced Inf values from division by zero. Once cleaned, this ranking surfaces Native 
American reservation counties (McKinley, NM; Big Horn, MT) and urban areas like the Bronx, 
NY, pointing to the disproportionate COVID toll on those communities.

No single metric tells the full story. The rankings diverge significantly depending on which 
dimension of burden is measured, which is exactly why looking at all three together matters.


# Question 2

```{r q2-load, message=FALSE, warning=FALSE}
# Load all three World Bank files
wbh      <- read_csv("worldbankhealth.csv")
wbh_class <- read_csv("WorldBankCountryClassClean.csv")
wbh_desc  <- read_csv("WorldBankHealth-VarDescClean.csv")

# Examine each file
head(wbh)
str(wbh)
head(wbh_class)
head(wbh_desc)
```

```{r q2-reshape, message=FALSE, warning=FALSE}
# Remove country_name and indicator_name — codes already capture this info
wbh2 <- wbh %>% select(-country_name, -indicator_name)

# Pivot wider: one row per country per year, each indicator becomes a column
wbh_wide <- wbh2 %>%
  pivot_wider(names_from = indicator_code, values_from = value)

# Examine the result
dim(wbh_wide)
head(wbh_wide[, 1:6])
```


```{r q2-select-cols, message=FALSE, warning=FALSE}
# Count NAs per column
na_counts <- apply(is.na(wbh_wide), MARGIN = 2, sum)

# Convert to data frame and sort by fewest NAs
na_df <- data.frame(column = names(na_counts), na_count = na_counts) %>%
  arrange(na_count)

head(na_df, 15)

# Select 11 columns with fewest NAs
top11_cols <- na_df$column[1:11]
wbh_final <- wbh_wide %>% select(all_of(top11_cols))

dim(wbh_final)
head(wbh_final)
```
Removing the two name columns keeps things cleaner since the codes already 
identify each country and indicator. After pivoting wider, each indicator gets 
its own column with one row per country per year. As expected, a lot of NAs show 
up since not every country reports every indicator every year. The 11 columns with 
the fewest NAs all turn out to be population and urbanization indicators, which 
makes sense as those tend to be the most consistently reported across countries.

# Question 3 - Box Plots

## Sepal width - base R
```{r q3-boxplot-base}
# Base R box plot of sepal width as instructed
boxplot(iris$Sepal.Width,
        main = "Sepal Width",
        ylab = "Width (cm)")
```
```{r q3-sepal-length}
# Vertical orientation is cleaner for a single variable
ggplot(iris, aes(y = Sepal.Length)) +
  geom_boxplot() +
  labs(title = "Sepal Length",
       y = "Length (cm)") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```


Sepal length across all iris observations has a median around 5.8 cm with the middle 
50% of values falling between roughly 5.1 and 6.4 cm. The distribution is fairly 
symmetric with no obvious outliers, and the whiskers extend from about 4.3 to 7.9 cm.

## Petal length vs class - ggplot
```{r q3-petal-length}
# Notched box plot of petal length by species, colored by species
ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot(notch = TRUE) +
  labs(title = "Petal Length by Species",
       x = "Species",
       y = "Petal Length (cm)") +
  theme(legend.position = "none")
```


Petal length differs dramatically across the three species. Setosa has by far the 
shortest petals (median around 1.5 cm) with very little variation, while versicolor 
sits in the middle (median around 4.3 cm) and virginica has the longest petals 
(median around 5.6 cm). The notches on all three boxes do not overlap, which 
indicates the medians are statistically significantly different from each other. 
Setosa also shows two outliers, and its very tight IQR suggests it is the most 
consistent of the three species in petal length.

## Petal width vs class - strip, jitter and violin
```{r q3-petal-width}
# Strip plot
ggplot(iris, aes(x = Species, y = Petal.Width, color = Species)) +
  geom_jitter(width = 0.2) +
  labs(title = "Petal Width by Species (Strip/Jitter)",
       x = "Species", y = "Petal Width (cm)") +
  theme(legend.position = "none")

# Violin plot
ggplot(iris, aes(x = Species, y = Petal.Width, fill = Species)) +
  geom_violin() +
  labs(title = "Petal Width by Species (Violin)",
       x = "Species", y = "Petal Width (cm)") +
  theme(legend.position = "none")
```


The jitter and violin plots tell the same story as the petal length analysis. Setosa 
stands out immediately with petals clustered tightly at around 0.2 cm — the jitter 
plot especially shows just how little variation there is across all 50 observations. 
Versicolor sits in the middle with values centered around 1.3 cm, and virginica has 
the widest petals, spreading mostly between 1.5 and 2.5 cm. The violin shapes are 
worth noting too: setosa looks almost flat and pancake-like because nearly all its 
values land at the same point, while virginica stretches tall showing more range in 
the data.

# Question 4 - Histograms

## Sepal width - base R
```{r q4-hist-base}
# Base R frequency histogram of sepal width as instructed
hist(iris$Sepal.Width,
     main = "Sepal Width Distribution",
     xlab = "Sepal Width (cm)",
     col  = "steelblue")
```


Sepal width is centered around 3.0 cm and has a pretty classic bell shape, ranging 
from about 2.0 to 4.4 cm. The tallest bar lands right at 3.0 with around 35 
observations. There is a slight right skew with a longer tail on the upper end, 
but overall the distribution is fairly symmetric.

## Sepal length - ggplot frequency histogram + polygon
```{r q4-sepal-length-hist}
# Part 1: Frequency histogram
# binwidth = 0.2 is the smallest that avoids empty bins
ggplot(iris, aes(x = Sepal.Length)) +
  geom_histogram(binwidth = 0.2, fill = "steelblue", color = "white") +
  labs(title = "Sepal Length Frequency Histogram",
       x = "Sepal Length (cm)",
       y = "Count")

# Part 2: Frequency polygon overlaid on histogram
ggplot(iris, aes(x = Sepal.Length)) +
  geom_histogram(binwidth = 0.2, fill = "steelblue", color = "white") +
  geom_freqpoly(binwidth = 0.2, color = "red", linewidth = 1) +
  labs(title = "Sepal Length with Frequency Polygon",
       x = "Sepal Length (cm)",
       y = "Count")
```


Sepal length ranges from about 4.3 to 7.9 cm with the tallest bar landing at 
5.0 cm. Rather than one clean peak, the distribution has two humps, which makes 
sense given that the three species have quite different sepal lengths pulling the 
overall shape apart. The frequency polygon traces that pattern nicely, making the 
rises and dips between bins easier to follow at a glance.

## Petal length - density histogram + kernel density
```{r q4-petal-length-density}
# Part 1: Density histogram
# binwidth = 0.2 is the smallest that avoids empty bins
ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 0.2, fill = "steelblue", color = "white") +
  labs(title = "Petal Length Density Histogram",
       x = "Petal Length (cm)",
       y = "Density")

# Part 2: Kernel density estimator overlaid
ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 0.2, fill = "steelblue", color = "white") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Petal Length Density with KDE",
       x = "Petal Length (cm)",
       y = "Density")
```


Petal length has a very clear two-humped shape, a tall narrow spike around 1.0 to 
1.5 cm and a broader second hump stretching from about 3.5 to 6.5 cm. That gap in 
the middle is setosa completely separated from the other two species. The KDE line 
smooths over the bin-by-bin noise and makes the bimodal structure even more obvious, 
though it undersells the height of the setosa spike since it is spreading the density 
across a wider range than the actual data.

## Petal width by class - faceted histogram + polygon
```{r q4-petal-width-facet}
# Part 1: Faceted frequency histogram, vertical orientation
ggplot(iris, aes(x = Petal.Width, fill = Species)) +
  geom_histogram(binwidth = 0.2, color = "white") +
  facet_wrap(~ Species, ncol = 1) +
  labs(title = "Petal Width by Species (Faceted Histogram)",
       x = "Petal Width (cm)",
       y = "Count") +
  theme(legend.position = "none")

# Part 2: Faceted frequency polygon
ggplot(iris, aes(x = Petal.Width, color = Species)) +
  geom_freqpoly(binwidth = 0.2, linewidth = 1) +
  facet_wrap(~ Species, ncol = 1) +
  labs(title = "Petal Width by Species (Faceted Frequency Polygon)",
       x = "Petal Width (cm)",
       y = "Count") +
  theme(legend.position = "none")
```


Breaking petal width out by species makes the differences really obvious. Setosa 
spikes sharply right at 0.2 cm and drops off almost immediately, confirming how 
tightly packed its values are. Versicolor has a much smoother, wider curve centered 
around 1.3 cm with a fairly even spread on both sides. Virginica is the most spread 
out of the three, with values running from about 1.4 to 2.5 cm and a flatter peak 
around 2.0 cm. The faceted layout makes it easy to compare the shapes side by side 
without the species overlapping and cluttering each other.