---
title: "Homework 2 - Data in R"
author: "Jules Pellin"
date: "4/23/2026"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r clear-env}
# Clear all previously defined objects from the environment before starting
rm(list = ls())
```

---

# Question 1 – Vectors

## (a) Crop Vector

```{r q1a}
# Create the crop vector with 9 character elements
crops <- c("apples", "wheat", "soybeans", "potatoes", "grapes",
           "cherries", "almonds", "hay", "hops")

print(crops)
```

## (b) Address Crop Vector

```{r q1b-replace}
# Replace soybeans (position 3) with oranges using index assignment
crops[3] <- "oranges"

print(crops)
```

```{r q1b-index-subset}
# Use indexing to subset only Washington major crops
# Washington major crops from the vector: apples (1), wheat (2), potatoes (4),
# grapes (5), cherries (6), hay (8), hops (9)
wa_crops_index <- crops[c(1, 2, 4, 5, 6, 8, 9)]

print(wa_crops_index)
```

```{r q1b-logical-subset}
# Use a logical vector to create the same Washington major crops subset
# TRUE for WA crops, FALSE for non-WA crops (oranges and almonds)
wa_logical <- c(TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE)

wa_crops_logical <- crops[wa_logical]

print(wa_crops_logical)
```

Both methods produce the same Washington major crop subset. The indexing method selects elements by their position number, while the logical method uses a TRUE/FALSE mask of the same length as the vector. They are just two different ways to get to the same result.

## (c) Reorder WA Crop Vector

```{r q1c-save}
# Save the Washington major crop vector from the indexing method
wa_crops <- wa_crops_index

print(wa_crops)
```

```{r q1c-reorder}
# Reorder to match the target output:
# "apples" "grapes" "cherries" "potatoes" "hops" "wheat" "hay"
# Current positions: apples=1, wheat=2, potatoes=3, grapes=4, cherries=5, hay=6, hops=7
wa_crops_reordered <- wa_crops[c(1, 4, 5, 3, 7, 2, 6)]

print(wa_crops_reordered)
```

```{r q1c-sort}
# Use order() to sort the Washington major crop vector alphabetically
wa_crops_sorted <- wa_crops[order(wa_crops)]

print(wa_crops_sorted)
```

order() returns the index positions that would sort the vector, and we use those positions to reorder it alphabetically. This is different from sort() which returns the sorted values directly. order() gives us the indices instead, which is more flexible when we need to apply the same ordering to multiple objects.

---

# Question 2 – Matrices

```{r q2-create-matrix}
# Create the base 4x5 matrix filled row by row with values 1 to 20
mat <- matrix(1:20, nrow = 4, ncol = 5, byrow = TRUE)

print(mat)
```

## (a) Address Matrix with Positive Indices

```{r q2a-subset1}
# Extract rows 2 to 4 and columns 3 to 4 to get the 3x2 subset
mat[2:4, 3:4]
```

```{r q2a-subset2}
# Extract row 2, columns 1, 3, and 5 to get the vector 6, 8, 10
mat[2, c(1, 3, 5)]
```

Positive indexing selects exactly the rows and columns we specify. A range like 2:4 is used for consecutive rows, while c(1, 3, 5) lets us pick columns that are not next to each other.

## (b) Subset Matrix with Negative Indices

```{r q2b-subset1}
# Remove row 3 using a negative index to get the 3x5 subset
mat[-3, ]
```

```{r q2b-subset2}
# Remove row 2 and columns 3 and 5 to get the 3x3 subset
mat[-2, c(-3, -5)]
```

Negative indexing excludes the specified rows or columns rather than selecting them. This is more efficient when we want to drop just one or two positions from a larger object.

## (c) Replace Values in Matrix

```{r q2c-replace-88}
# Create a duplicate matrix so the original stays unaltered
mat_c1 <- mat

# Replace values 9, 14, and 19 (all in column 4, rows 2 to 4) with 88
mat_c1[2:4, 4] <- 88

print(mat_c1)
```

```{r q2c-replace-individual}
# Create a fresh duplicate from the original matrix
mat_c2 <- mat

# Replace 9, 14, and 19 with 23, 24, 25 respectively
mat_c2[2:4, 4] <- c(23, 24, 25)

print(mat_c2)
```

```{r q2c-remove-8}
# Create a fresh duplicate from the original matrix
mat_c3 <- mat

# Remove the value 8 (row 2, column 3) by setting it to NA
mat_c3[2, 3] <- NA

print(mat_c3)
```

Each replacement starts from a fresh copy of the original matrix so previous changes don't carry over. Replacing with a single value (88) applies it to all three positions at once, while replacing with c(23, 24, 25) maps each value to its corresponding position. Setting a value to NA is how we remove it in R, which marks the cell as missing rather than deleting the structure.

---

# Question 3 – Simple Lists

## list1

```{r q3-list1-create}
# Create list1 with three elements: a string, a numeric vector, and a matrix
list1 <- list(
  "Soils 502",
  c(1, 3, 4, 5),
  matrix(1:6, nrow = 3, ncol = 2)
)

print(list1)
```

```{r q3-list1-extract-matrix}
# Extract the matrix (element 3) using double brackets
list1[[3]]
```

```{r q3-list1-extract-sublist}
# Extract elements 1 and 3 together as a sub-list using single brackets
list1[c(1, 3)]
```

Double brackets [[]] return the actual content of a single element, so in this case we get the matrix itself. Single brackets [] return a sub-list that still wraps the selected elements. The key distinction is that [[]] unwraps the element while [] keeps it packaged as a list.

## (b) list2

```{r q3-list2-create}
# Create list2 with Serena Williams data
# name is a nested list with Last and First as sub-elements
# grand.slam.wins is also a nested list, one element per tournament
list2 <- list(
  name = list(
    Last  = "Williams",
    First = "Serena"
  ),
  age    = 36,
  record = c(wins = 801, losses = 136),
  grand.slam.wins = list(
    Australian.Open = c(2003, 2005, 2007, 2009, 2010, 2015, 2017),
    French.Open     = c(2002, 2013, 2015),
    Wimbledon       = c(2002, 2003, 2009, 2010, 2012, 2015, 2016),
    US.Open         = c(1999, 2002, 2008, 2012, 2013, 2014)
  )
)

print(list2)
```

**Why aren't grand slam wins structured as a data frame or matrix?**

A matrix can only hold one data type, so mixing tournament names with year values would force everything to become a character string and we would lose the ability to do any numeric calculations. A data frame could technically work but each tournament has a different number of wins, with Australian Open having 7, French Open having 3, Wimbledon having 7, and US Open having 6. Since a data frame requires equal-length columns, it would not fit cleanly either. A list is the right structure here because each element can have a different length and type without any coercion.

---

# Question 4 – cbind, rbind

```{r q4-create-objects}
# Create the five required data objects

# Numeric vector with 4 elements
v1 <- c(3, 5, 6, 12)

# Logical vector with 4 elements
v2 <- c(TRUE, FALSE, TRUE, TRUE)

# Character vector with 5 elements
v3 <- c("A", "B", "C", "D", "E")

# 4x5 numeric matrix
m1 <- matrix(c(2, 14, 26, 5, 17,
               5, 17, 29, 8, 20,
               8, 20, 32, 11, 23,
               11, 23, 2, 14, 26),
             nrow = 4, ncol = 5, byrow = TRUE)

# 2x4 numeric matrix
m2 <- matrix(c(1, -1, 1, 0,
               0,  0, -1, 1),
             nrow = 2, ncol = 4, byrow = TRUE)

print(v1); print(v2); print(v3); print(m1); print(m2)
```

## (a) cbind

```{r q4a-cbind1}
# cbind v1 and v2: logical TRUE/FALSE coerces to numeric 1/0
cbind(v1, v2)
```

```{r q4a-cbind2}
# cbind v3 and v2: all values become character strings since v3 is character
cbind(v3, v2)
```

```{r q4a-cbind3}
# cbind m1 and v1: v1 becomes a 6th column added to the right of m1
cbind(m1, v1)
```

cbind() places objects side by side as columns and the order of arguments determines which column appears first. When types are mixed, R converts everything to the most flexible type, so logical combined with numeric becomes 1 and 0, and character combined with anything becomes character strings.

## (b) rbind

```{r q4b-rbind1}
# rbind the first 4 columns of m1 with m2 (both have 4 columns)
rbind(m1[, 1:4], m2)
```

```{r q4b-rbind2}
# rbind v1 and m2: v1 becomes the top row, m2 rows follow beneath
rbind(v1, m2)
```

```{r q4b-rbind3}
# rbind v1, v2, and v3: all values become character since v3 is character
# v1 and v2 each have 4 elements and v3 has 5, so R recycles v1 and v2
rbind(v1, v2, v3)
```

rbind() stacks objects on top of each other as rows and the first argument becomes the top rows in the output. When a shorter vector is combined with a longer one, R recycles the shorter vector to match the length of the longest. This is why v1 shows "3" again in the 5th column position.